Creep and fracture of a protein gel under stress

### 2014-04-23 creep/creep_yaourt.tex
Much more care in the use of "brittle" (disapear from the title) yield stress fluids in intro Move the discussion on reversibility to the discussion. Reduce out of ROI fracture influence Aknowledge Mikko Alava Tabuteau:2009, Skrzeszewska:2010 SupplFig3 et 4

 \title{Creep and brittle failure of a protein gel under stress}

@@

\begin{abstract}
Biomaterials such as protein or polysaccharide gels are known to behave qualitatively as soft solids and to rupture under an external load. Combining optical and ultrasonic imaging to shear rheology we show that the failure scenario of a protein gel is typical of brittle solids: after a power-law creep regime fully accounted for by linear viscoelasticity and homogeneous deformation, fractures nucleate and grow logarithmically perpendicularly to shear up to sudden rupture. A single equation accounting for those two successive processes nicely captures the full rheological response. The failure time follows a decreasing power-law with the applied shear stress strongly reminiscent of the Basquin law of fatigue for solids. These results are in excellent agreement with recent fiber-bundle models that include damage accumulation on elastic fibers and exemplify protein gels as model brittle soft solids.
\end{abstract}

@@

%Under a constant external load, a solid deforms and may eventually fail. The resulting slow deformation, referred to as {\it creep}, has been investigated over a century on solid materials \cite{Andrade:1910,Nechad:2005,Miguel:2002} but only a handful of papers have dealt with creep on soft solids, i.e. gels made through the reversible association of biopolymers such as proteins or polysaccharides \cite{Higgs:1990,vanVliet:1995}. These materials, whose physical properties play a key role not only in food science, from set-style yogurt or cheese to jelly-based deserts \cite{Mezzenga:2005} but also in biomechanics, biochemistry and microbiology \cite{Viovy:2000,Grinnell:2010}, behave as elastic solids under small deformations but display remarkable nonlinear behavior featuring fractures prior to irreversible rupture \cite{Bonn:1998,Baumberger:2006,Daniels:2007,Brenner:2013}. Irreversibility directly stems from the existence of an external control parameter, namely temperature (resp. pH) in the case of thermoreversible gels (resp. acid-induced gels), which makes the system impossible to shear-rejuvenate after failure, contrary to other soft glassy materials such as emulsions, colloidal gels and glasses \cite{Cloitre:2000,Divoux:2012,Siebenburger:2012}. While great effort has been devoted to designing protein- and polysaccharide-based soft solids with specific properties and textures at rest \cite{Dickinson:2006,Gibaud:2012}, their nonlinear mechanical behavior under a shear stress has only been partially addressed \cite{vanVliet:1995} and several simple issues remain unexplored such as the spatially resolved scenario for creep and fracture or the physical relevance of the analogy with the failure of hard, brittle solids.

Biogels formed through the self-association of polysaccharide coils, collagen, actin filaments or attractive globular proteins play a major role in biochemistry and microbiology \cite{Viovy:2000}, biological networks and cell mechanics \cite{Stricker:2010} as well as in food science \cite{Mezzenga:2005}. These biomaterials all behave as elastic solids under small deformations but display remarkable nonlinear behavior generally featuring stress- or strain-stiffening \cite{Gardel:2004} and fractures prior to irreversible rupture \cite{Bonn:1998,Baumberger:2006}. Irreversibility stems from the existence of an external control parameter, e.g. temperature or pH in the case of thermoreversible or acid-induced gels respectively, which makes such biogels impossible to shear-rejuvenate after failure, contrary to other soft glassy materials such as emulsions, colloidal gels and glasses \cite{Cloitre:2000,Divoux:2012,Siebenburger:2012}. So far huge effort has been devoted to designing protein gels with specific properties and textures at rest \cite{Dickinson:2006,Gibaud:2012}. However, their mechanical behavior deep into the nonlinear regime has only been partially addressed \cite{vanVliet:1995,Pouzot:2006} and several fundamental issues remain unexplored such as the spatially resolved rupture scenario or the physical relevance of the analogy with brittle failure in hard solids.

%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=7cm]{Fig1.eps}
\caption{(color online) Strain response $\gamma(t)$ in a 4\%~wt. casein gel acidified with 1\%~wt. GDL for an imposed shear stress $\sigma=200$ (\textcolor{red!25!black}{$\bullet$}), 300 (\textcolor{red!50!black}{$\blacktriangledown$}), 400 (\textcolor{red!75!black}{$\blacksquare$}), 550 (\textcolor{red}{$\blacktriangle$}), and 1000~Pa (\textcolor{orange!50!red}{$\blacklozenge$}) from right to left. The gap width is 1~mm. Gray dashes show $\gamma=1$. Inset: failure time $\tau_f$ vs $\sigma$. The red line is the best power-law fit $\tau_f=A\sigma^{-\beta}$ with $\beta=5.45\pm 0.05$ and $A=40.6\pm 0.5$~s.Pa$^\beta$.
\label{fig1}}
\end{figure} 
%%%%%%%%%%%%

In this Letter we report on stress-induced fracture in protein gels by means of creep experiments coupled to optical and ultrasonic imaging. Gels formed by slow acidification of a sodium caseinate solution \cite{Lucey:1998} display fractures under large strain at fixed low pH values \cite{vanVliet:1995}, which makes them perfect candidates to quantify brittleness in soft solids and tackle the above-mentioned issues. We demonstrate that under an external load, these casein gels display brittle-like failure that results from two successive physical processes: a reversible primary creep regime characterized by a homogeneous strain field and the irreversible nucleation and fracture growth leading to gel failure. Our results are in full agreement with the predictions of some recent fiber-bundle models and hint to universal features of brittle failure common to both soft and hard solids.

@@

%%%%%%%%%%%%
\begin{figure*}[t]
\centering
\includegraphics[width=17.5cm,clip]{Fig3.eps}
\caption{(color online) Ultrasonic images (left) and direct visualization of the Couette cell (right) recorded simultaneously at various times in the primary [(a)~$t/\tau_f=0.02$], secondary [(b)~$t/\tau_f=0.57\simeq\taum/\tau_f$], and tertiary  [(c) $t/\tau_f=0.93$ and (d) 0.99] creep regimes. Ultrasonic images are velocity maps $v(r,z,t)$ computed by averaging over 4~s and coded using the linear color levels shown below the images. Their position relative to the picture of the cell reflects the actual arrangement of the ultrasonic probe along the vertical direction $z$. The gap width is 2~mm and $r$ is the radial distance to the inner rotating cylinder. Experiment performed on a 4\%~wt. casein gel seeded with 3\%~wt. polyamide spheres and acidified with 1\%~wt. GDL under $\sigma=300$~Pa. See also Movie~2 in the Supplemental Material.
\label{fig3}}
\end{figure*} 
%%%%%%%%%%%%

As seen in the inset of Fig.~\ref{fig1}, the failure time $\tau_f$ sharply decreases as a power-law of $\sigma$ with an exponent $\beta\simeq 5.5$. $\tau_f$ further allows us to rescale all the shear rate data $\gp(t)$ onto the single master curve of Fig.~\ref{fig2}(a) by plotting $\gp/\gp_{\rm min}$ vs $t/\tau_f$, where $\gp_{\rm min}$ is the minimum shear rate reached at a time $\tau_{\rm min}$ [see also Fig.~\ref{suppfig2} in the Supplemental Material for the original $\gp(t)$ data]. On about four decades for $t\lesssim 0.1\tau_f$, the shear rate decreases as a power-law $\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.02$. This is strongly reminiscent of the {\it primary creep} observed in solids and referred to as Andrade creep \cite{Andrade:1910,Miguel:2002,Nechad:2005}. Interestingly, here, this scaling can be fully accounted for by linear viscoelasticity without invoking any irreversible plasticity. Indeed casein gels display a power-law rheology $G'(f)\sim G''(f)\sim f^{0.15}$ [inset of Fig.~\ref{fig2}(a)], which corresponds to a compliance $J(t)\equiv\gamma(t)/\sigma\sim t^{0.15}$ in the linear deformation regime \cite{Tschoegl:1989}. This nicely corresponds to the observed $\gp(t)\sim t^{-0.85}$ and suggests that primary creep is purely reversible without unrecoverable plastic events. Homogeneity and linear response are further confirmed, at least up to the available spatial ($\sim 10~\mu$m) and temporal ($\sim 1$~s) resolutions, by direct visualization which shows no sign of crack or fracture [Fig.~\ref{fig3}(a)] and by ultrasonic imaging which reveals velocity and strain fields averaged over the vertical direction $z$ that linearly decrease with the radial position $r$ within the gap [Fig.~\ref{fig4}(a-b)] with insignificant $z$-dependence [Fig.~\ref{fig3}(a, left)] and no slippage at the Plexiglas walls.

For $0.1\lesssim t/\tau_f\lesssim 0.9$, $\gp(t)$ departs from power-law behavior and goes through a minimum value at time $\tau_{\rm min}= (0.56\pm0.04)\tau_f$ independently of the applied stress [Fig.~\ref{fig2}(b) and inset] as similarly reported for metals \cite{Sundararajan:1989}, solid composite materials \cite{Nechad:2005} and fiber-bundles models (FBMs) \cite{Kovacs:2008,Jagla:2011}. This linearity between $\tau_{\rm min}$ and $\tau_f$, also known as the Monkman-Grant relation \cite{Monkman:1956}, allows one to ``predict'' the failure time from the intermediate-time response. During this {\it secondary creep} regime, regularly-spaced cracks nucleate from the top and bottom edges of the Couette cell and start growing perpendicular to the applied stress [Fig.~\ref{fig3}(b), see also Movie~2 in the Supplemental Material]. These macroscopic fractures are invaded with water expelled from the surrounding gel matrix. Although fractures have not yet entered the ultrasonic region of interest, velocity maps become heterogeneous along the $z$ direction [Fig.~\ref{fig3}(b, left)] and display intermittent fluctuations [Fig~\ref{fig4}(c,d)]. The level of these fluctuations, which can also be seen on the global response for the same applied stress [\textcolor{red!50!black}{$\blacktriangledown$} in Fig.~\ref{fig2}(b)], is poorly reproducible and stress-dependent. We propose that such intermittency arises from microscopic cracks ahead of the fracture tip. Since heterogeneity is much smaller along the radial direction [compare red and black lines in Fig~\ref{fig4}(d)], these ``precursors'' either extend across the full 2-mm gap or induce long-range displacements that span the gap. In any case, such noisy features are typical of slow crack growth in solids \cite{Vanel:2009,Kun:2007} and more generally of avalanche behavior \cite{Sethna:2001}.

@@

%%%%%%%%%%%%
\begin{figure*}
\centering
\includegraphics[width=17.5cm,clip]{Fig4.eps}
\caption{(color online) (a)~Local velocity $\langle v(r,z,t)\rangle_z$ and (b)~local strain field $\langle\gl(r,z,t)\rangle_z$ averaged over the vertical direction $z$ at various times during primary creep: $t/\tau_f=1.9\,10^{-3}$ ($\bullet$), $1.7\,10^{-2}$ (\textcolor{blue}{$\blacktriangledown$}), and 0.15 (\textcolor{red}{$\blacksquare$}). Solid lines are linear profiles. (c)~Spatiotemporal diagram of the local velocity $\langle v(r,z,t)\rangle_r$ averaged over the radial direction $r$ and plotted in linear color levels as a function of $z$ and $t/\tau_f$. (d)~Standard deviation $\delta_z v(t)$ of $\langle v(r,z,t)\rangle_r$ taken over the vertical direction $z$ (thick black line) together with corresponding standard deviation $\delta_r v(t)$ computed over the radial direction $r$ on the $z$-average $\langle v(r,z,t)\rangle_z$ (thin red line). (e)~Fracture length $\ell(t)$ vs $(\tau_f-t)/\tau_f$ as inferred from direct visualization ($\bullet$, average over 6 different fractures, error bars show the standard deviation) and from ultrasonic imaging ($\circ$) and normalized by the height $H$ of the Couette cell. Gray dots show the visualization data for the longest fracture which leads to the failure of the sample at $\tau_f$. Red lines are the best fits $\ell(t)=a+b\log(1-t/\tau_f)$ to the visualization data. Same experiment as in Fig.~\ref{fig3} and Supplemental Movie~2.
\label{fig4}}
\end{figure*} 
%%%%%%%%%%%%

Finally, we emphasize that the (reversible) Andrade-like creep and the (irreversible) crack growth are two  physical processes that effectively superimpose to yield the global rheological response. Indeed, as seen from the yellow line in Fig.~\ref{fig2}, the master curve $\gp(t)/\gp_{\rm min}$ vs $t/\tau_f$ is perfectly fitted by:
%the master curve $\tilde{\gp}(t)=\gp(t)/\gp_{\rm min}$ vs $\tilde{t}=t/\tau_f$ is perfectly fitted by:
\begin{equation}
\frac{\gp(t)}{\gp_{\rm min}}=\lambda \left(\frac{t}{\tau_f}\right)^{-\alpha}+ \frac{\mu}{1-t/\tau_f}\,,
%\tilde{\gp}(\tilde{t})=\lambda \tilde{t}^{-\alpha}+ \frac{\mu}{1-\tilde{t}}\,,
\label{e.fit}
\end{equation}
with only two adjustable parameters $\lambda$ and $\mu$ once $\alpha=0.85$ is fixed. The remarkable collapse of the whole data set to such a simple equation allows us to interpret the secondary creep regime as a mere crossover from creep to crack growth.

Let us now summarize and discuss the most prominent results of this Letter. First we have shown that casein gels display a remarkable failure scenario similar to that of brittle solids and characterized by the same three successive creep regimes. Here, the primary creep and its power-law exponent [$\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.02$] are fully accounted for by linear viscoelasticity and local measurements consistently display a homogeneous bulk deformation. Such a link between creep and viscoelasticity is shared by other biopolymer gels with power-law rheology \cite{Gobeaux:2010,Jaishankar:2013} as well as hard-sphere-like colloidal glasses \cite{Siebenburger:2012}. Protein gels thus appear to remain purely elastic and reversible throughout primary creep, a hallmark of brittle materials. By contrast primary creep in solids is irreversible and has been attributed to dislocation interactions in the case of crystalline materials \cite{Miguel:2002} or to localized plastic rearrangements in amorphous materials \cite{Rosti:2010}.

The logarithmic fracture growth in the tertiary creep regime constitutes our second important result. Such an evolution is also commonly reported in disordered solid materials displaying brittle rupture and interpreted in the framework of Griffith-like models based on global or local energy barriers \cite{Vanel:2009}. However these approaches all predict exponential scalings for $\tau_f(\sigma)$ while our data are best fitted by the decreasing power-law $\tau_f \sim \sigma^{-\beta}$ with $\beta=5.45\pm 0.05$ [inset of Fig.~\ref{fig1}]. This last key result suggests that thermally activated crack growth is not relevant to the present soft solid. Rather the power-law scaling is strikingly reminiscent of the Basquin law of fatigue found for a variety of heterogeneous or cellular materials under cyclic deformation \cite{Kun:2007,Basquin:1910,Kohout:2000}. Basquin law has also been recently predicted by FBMs that combine elastic fibers with a local yield strain and take into account damage accumulation \cite{Kun:2007,Halasz:2012}. Interestingly assuming damage accumulation to be proportional to $\sigma^\gamma$ directly leads to Basquin law with $\beta=\gamma$ and large values of $\gamma$, typically larger than 5 as found here for $\beta$, lead to macroscopic cracks due to the simultaneous rupture of a large number of fibers \cite{Halasz:2012}. More generally FBMs under compressive or elongational load predict three successive creep regimes exactly alike Fig.~\ref{fig2} with similar proportionality between $\tau_{\rm min}$ and $\tau_f$ and finite-time singularity \cite{Nechad:2005,Jagla:2011}. This highlights the relevance of FBMs in the context of creep in protein gels, whose microstructure indeed appears to be formed of strands \cite{Kalab:1983,Roefs:1990}, but also urges to study FBMs in shear geometries to check whether they would be able to predict fracture growth as observed here.

To conclude, the present time- and space-resolved study shows that protein gels qualify as model, brittle soft solids. Prompted by the remarkable simplicity of Eq.~(\ref{e.fit}) that encompasses Andrade creep, the Monkman-Grant relation and finite-time singularity within a single equation, future modeling will undoubtedly focus on the microscopic ingredients needed to predict quantitatively the Monkman-Grant prefactor and the Basquin exponent $\beta$. The next experimental step consists in a statistical study of the fluctuations associated to crack nucleation and growth as well as a systematic investigation of other systems in order to check for universality in the irreversible creep rupture of soft solids in general and of biogels such as actin, alginate or agar gels in particular. Such a study is expected to have important implications in understanding the behavior of biomaterials under extreme stress conditions.

%which separates into three successive phases: (i)~a primary (or Andrade-like) creep regime during which the strain field is homogeneous, (ii)~a secondary creep during which fractures nucleate and grow logarithmically, and (iii)~a finite-time singularity concomittant with the sudden failure of the gel. The failure time displays a non-trivial power-law scaling with the applied shear-stress that contrasts with the exponential scalings usually reported for solid brittle materials \cite{Vanel:2009}.

% within the spatial resolution of a few microns, the strain field is homogeneous [Fig~\ref{fig4}~(b)]. If plastic events take place in this primary regime, they occur at smaller scale. However, given the agreement between the Andrade exponent and the frequency scaling of the linear elastic modulus discussed above, it is more likely that no or a negligible amount of plastic events take place during this regime, corroborating the optical and ultrasonic measurements.

%Compare to the scalings observed in YSF - in particular microgels display a similar power-law scaling, except that the yield point appears explicitely in the scaling as the critical point in a dynamical phase transition \cite{Divoux:2012}. Propose an explanation or at least raise a point. \cite{Chaudhuri:2013}

%FBM that rely on elastic fibers with a local yield strain (or stress) and introduce a non-linear Eyring rheology or a local coupling lead to a different scaling law than the one we find here for the failure time with the applied stress. In their law appears a ``yield stress" \cite{Kun:2003,Nechad:2005,Kovacs:2008,Jagla:2011} that is not present here.

%different from other soft materials such as transient-network systems \cite{Ligoure:2013}, because truly irreversible. The pH is key to the irreversibility and plays the role of an additional control parameter that, when maintained at values lower than the isoelectric point, prevent the system from healing.

%Comparatively, Yield Stress Fluids (YSF) that display a solid-like behavior at rest, either because of a jammed microstructure or attractive interactions \cite{Sciortino:2005,Bonnecaze:2010}, but do not present such external control parameter, display a similar rheological signature during their stress-induced yielding. Indeed the S-shaped curved pictured in figure~\ref{fig2}~(a) has been also reported in \cite{Bauer:2006,Gibaud:2010,Divoux:2011,Siebenburger:2012,Grenard:2013}. However, their yielding scenario is fundamentally different. Local measurements conducted on both type of soft glassy materials have revealed that no fractures take place in the bulk: the microstructure rather reorganizes and flows in a ductile way. Our study reveals that acid-induced gels behave elastically up to the rupture point and that fractures develop and grow logarithmically, perpendicularly to the shear direction prior to an irreversible failure which makes casein gels model {\it brittle soft solids}.

%and yield stress fluids (YSFs) \cite{Bauer:2006,Divoux:2011,Siebenburger:2012,Grenard:2013} that usually exhibit a power-law creep with smaller exponents (0.3 $\lesssim \alpha \lesssim 0.7$) which are not directly related to linear rheology. Besides, the linear behavior of YSF generally differs from a simple power-law scaling, while in the case of solids

\begin{acknowledgments}
The authors thank T.~Gibaud, S.~Lindstr\"om and N.~Taberlet for fruitful discussions, M.~Kal\'ab for kindly providing us with unobtainable references, and A.~Parker at Firmenich for precious advice and for providing the casein and GDL. This work was funded by the Institut Universitaire de France and by the European Research Council under the European Union's Seventh Framework Programme (FP7/2007-2013) / ERC grant agreement No.~258803. 
\end{acknowledgments}

@@

\bibitem{Divoux:2012} T. Divoux et al., Soft Matter {\bf 8}, 4151--4164 (2012).

\bibitem{Siebenburger:2012} Siebenb\"urger, M. Ballauff and T. Voigtmann, Phys. Rev.
Lett., {\bf 108}, 255701 (2012).

>>>>

 \title{Creep and fracture of a protein gel under stress}

@@

\begin{abstract}
Biomaterials such as protein or polysaccharide gels are known to behave qualitatively as soft solids and to rupture under an external load. Combining optical and ultrasonic imaging to shear rheology we show that the failure scenario of a protein gel is reminiscent of brittle solids: after a primary creep regime characterized by a macroscopically homogeneous deformation and a power-law behavior which exponent is fully accounted for by linear viscoelasticity, fractures nucleate and grow logarithmically perpendicularly to shear, up to the sudden rupture of the gel. A single equation accounting for those two successive processes nicely captures the full rheological response. The failure time follows a decreasing power-law with the applied shear stress, similar to the Basquin law of fatigue for solids. These results are in excellent agreement with recent fiber-bundle models that include damage accumulation on elastic fibers and exemplify protein gels as model, brittle-like soft solids.
\end{abstract}

@@

Biogels formed through the self-association of polysaccharide coils, collagen, actin filaments or attractive globular proteins play a major role in biochemistry and microbiology \cite{Viovy:2000}, biological networks and cell mechanics \cite{Stricker:2010} as well as in food science \cite{Mezzenga:2005}. These biomaterials all behave as elastic solids under small deformations but display remarkable nonlinear behavior generally featuring stress- or strain-stiffening \cite{Gardel:2004} and fractures prior to irreversible rupture \cite{Bonn:1998,Baumberger:2006}. Irreversibility stems from the existence of an external control parameter, e.g. temperature or pH in the case of thermoreversible or acid-induced gels respectively. This makes such biogels fundamentally different from other soft glassy materials such as emulsions, colloidal gels and glasses that can be rejuvenated by shear \cite{Cloitre:2000,Caton:2008,Divoux:2011,Siebenburger:2012} or transient networks where fractures spontaneously heal \cite{Tabuteau:2009,Skrzeszewska:2010}. So far, huge effort has been devoted to the design of protein gels with specific properties and textures at rest \cite{Dickinson:2006,Gibaud:2012}. However, their mechanical behavior deep into the nonlinear regime has only been partially addressed \cite{vanVliet:1995,Pouzot:2006} and several fundamental issues remain unexplored such as the spatially resolved rupture scenario or the physical relevance of the analogy with brittle failure in hard solids.

%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=7cm]{Fig1.eps}
\caption{(color online) Strain response $\gamma(t)$ in a 4\%~wt. casein gel acidified with 1\%~wt. GDL for an imposed shear stress $\sigma=200$ (\textcolor{red!25!black}{$\bullet$}), 300 (\textcolor{red!50!black}{$\blacktriangledown$}), 400 (\textcolor{red!75!black}{$\blacksquare$}), 550 (\textcolor{red}{$\blacktriangle$}) and 1000~Pa (\textcolor{orange!50!red}{$\blacklozenge$}) from right to left. The gap width is 1~mm. Gray dashes show $\gamma=1$. Inset: failure time $\tau_f$ vs $\sigma$. The red line is the best power-law fit $\tau_f=A\sigma^{-\beta}$ with $\beta=5.45\pm 0.05$ and $A=40.6\pm 0.5$~s.Pa$^\beta$.
\label{fig1}}
\end{figure} 
%%%%%%%%%%%%

In this Letter we report on stress-induced fracture in protein gels by means of creep experiments coupled to optical and ultrasonic imaging. Gels formed by slow acidification of a sodium caseinate solution display fractures under large strain at fixed low pH values \cite{vanVliet:1995,Lucey:1998}, which makes them perfect candidates to quantify the rupture of soft solids and tackle the above-mentioned issues. We demonstrate that under an external load, these casein gels display brittle-like failure that results from two successive physical processes: (i) a primary creep regime characterized by a macroscopically homogeneous strain field where dissipation is dominated by viscous flow through the gel matrix and (ii) the irreversible nucleation and growth of fractures leading to gel failure. Our results are in full agreement with the predictions of some recent fiber-bundle models and hint to universal features of failure common to both soft and hard solids.

@@

%%%%%%%%%%%%
\begin{figure*}[t]
\centering
\includegraphics[width=17.5cm,clip]{Fig3.eps}
\caption{(color online) Ultrasonic images (left) and direct visualization of the Couette cell (right) recorded simultaneously at various times in the primary [(a)~$t/\tau_f=0.02$], secondary [(b)~$t/\tau_f=0.57\simeq\taum/\tau_f$] and tertiary  [(c) $t/\tau_f=0.93$ and (d) 0.99] creep regimes. Ultrasonic images are azimuthal velocity maps $v(r,z,t)$ computed by averaging over 4~s and coded using the linear color levels shown below the images. Their position relative to the picture of the cell reflects the actual arrangement of the ultrasonic probe along the vertical direction $z$. The gap width is 2~mm and $r$ is the radial distance to the inner rotating cylinder. Experiment performed on a 4\%~wt. casein gel seeded with 3\%~wt. polyamide spheres and acidified with 1\%~wt. GDL under $\sigma=300$~Pa. See also Movie~2 in the Supplemental Material.
\label{fig3}}
\end{figure*} 
%%%%%%%%%%%%

As seen in the inset of Fig.~\ref{fig1}, the failure time $\tau_f$ sharply decreases as a power-law of $\sigma$ with an exponent $\beta\simeq 5.5$. $\tau_f$ further allows us to rescale all the shear rate data $\gp(t)$ onto the single master curve of Fig.~\ref{fig2}(a) by plotting $\gp/\gp_{\rm min}$ vs $t/\tau_f$, where $\gp_{\rm min}$ is the minimum shear rate reached at a time $\tau_{\rm min}$ [see also Fig.~\ref{suppfig2} in the Supplemental Material for the original $\gp(t)$ data]. On about four decades for $t\lesssim 0.1\tau_f$, the shear rate decreases as a power-law $\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.04$. This is strongly reminiscent of the {\it primary creep} observed in solids and referred to as Andrade creep \cite{Andrade:1910,Miguel:2002,Nechad:2005,Rosti:2010}. Interestingly, here, the exponent can be inferred from linear viscoelasticity. Indeed casein gels display a power-law rheology $G'(f)\sim G''(f)\sim f^{0.15}$ [inset of Fig.~\ref{fig2}(a)], which corresponds to a compliance $J(t)\equiv\gamma(t)/\sigma\sim t^{0.15}$ in the linear deformation regime \cite{Tschoegl:1989}. This nicely corresponds to the observed $\gp(t)\sim t^{-0.85}$. Moreover ultrasonic imaging reveals velocity and strain fields averaged over the vertical direction $z$ that linearly decrease with the radial position $r$ within the gap [Fig.~\ref{fig4}(a-b)] with insignificant $z$-dependence [Fig.~\ref{fig3}(a, left)] and no slippage at the Plexiglas walls. Together with direct visualization [Fig.~\ref{fig3}(a)], these local measurements demonstrate that during primary creep there is no macroscopic strain localization or fracture, although we cannot rule out microscopic rearrangements below the available spatial ($\sim10~\mu$m) and temporal ($\sim1$~s) resolutions.

For $0.1\lesssim t/\tau_f\lesssim 0.9$, $\gp(t)$ departs from power-law behavior and goes through a minimum value at time $\tau_{\rm min}= (0.56\pm0.04)\tau_f$ independently of the applied stress [Fig.~\ref{fig2}(b) and inset] as similarly reported for metals \cite{Sundararajan:1989}, solid composite materials \cite{Nechad:2005} and fiber-bundle models (FBMs) \cite{Kovacs:2008,Jagla:2011}. This linearity between $\tau_{\rm min}$ and $\tau_f$, also known as the Monkman-Grant relation \cite{Monkman:1956}, allows one to ``predict'' the failure time from the intermediate-time response. During this {\it secondary creep} regime, regularly-spaced cracks nucleate from the top and bottom edges of the Couette cell and start growing perpendicular to the applied stress [Fig.~\ref{fig3}(b), see also Movie~2 in the Supplemental Material]. These macroscopic fractures are invaded with water expelled from the surrounding gel matrix. Although fractures have not yet entered the ultrasonic region of interest, velocity maps become heterogeneous along the $z$ direction [Fig.~\ref{fig3}(b, left)] and display intermittent fluctuations [Fig~\ref{fig4}(c,d)]. The level of these fluctuations, which can also be seen on the global response for the same applied stress [\textcolor{red!50!black}{$\blacktriangledown$} in Fig.~\ref{fig2}(b)], is poorly reproducible and stress-dependent. Such intermittency may arise from crack growth outside the region of interest which induces long-range elastic deformation propagating across the sample mainly along the vertical direction [see the lower level of fluctuations along the radial direction in Fig~\ref{fig4}(d)].

@@

%%%%%%%%%%%%
\begin{figure*}
\centering
\includegraphics[width=17.5cm,clip]{Fig4.eps}
\caption{(color online) (a)~Local velocity $\langle v(r,z,t)\rangle_z$ and (b)~local strain field $\langle\gl(r,z,t)\rangle_z$ averaged over the vertical direction $z$ at various times during primary creep: $t/\tau_f=1.9\,10^{-3}$ ($\bullet$), $1.7\,10^{-2}$ (\textcolor{blue}{$\blacktriangledown$}) and 0.15 (\textcolor{red}{$\blacksquare$}). Solid lines are linear profiles. (c)~Spatiotemporal diagram of the local velocity $\langle v(r,z,t)\rangle_r$ averaged over the radial direction $r$ and plotted in linear color levels as a function of $z$ and $t/\tau_f$. (d)~Standard deviation $\delta_z v(t)$ of $\langle v(r,z,t)\rangle_r$ taken over the vertical direction $z$ (thick black line) together with corresponding standard deviation $\delta_r v(t)$ computed over the radial direction $r$ on the $z$-average $\langle v(r,z,t)\rangle_z$ (thin red line). (e)~Fracture length $\ell(t)$ vs $(\tau_f-t)/\tau_f$ as inferred from direct visualization ($\bullet$, average over 6 different fractures, error bars show the standard deviation) and from ultrasonic imaging ($\circ$) and normalized by the height $H$ of the Couette cell. Gray dots show the visualization data for the longest fracture which leads to the failure of the sample at $\tau_f$. Red lines are the best fits $\ell(t)=a+b\log(1-t/\tau_f)$ to the visualization data. Same experiment as in Fig.~\ref{fig3} and Supplemental Movie~2.
\label{fig4}}
\end{figure*} 
%%%%%%%%%%%%

Finally, we emphasize that the Andrade-like creep and the crack growth are two  physical processes that effectively superimpose to yield the global rheological response. Indeed, as seen from the yellow line in Fig.~\ref{fig2}, the master curve $\gp(t)/\gp_{\rm min}$ vs $t/\tau_f$ is perfectly fitted by:
%the master curve $\tilde{\gp}(t)=\gp(t)/\gp_{\rm min}$ vs $\tilde{t}=t/\tau_f$ is perfectly fitted by:
\begin{equation}
\frac{\gp(t)}{\gp_{\rm min}}=\lambda \left(\frac{t}{\tau_f}\right)^{-\alpha}+ \frac{\mu}{1-t/\tau_f}\,,
%\tilde{\gp}(\tilde{t})=\lambda \tilde{t}^{-\alpha}+ \frac{\mu}{1-\tilde{t}}\,,
\label{e.fit}
\end{equation}
with only two adjustable parameters $\lambda$ and $\mu$ once $\alpha=0.85$ is fixed. The remarkable collapse of the whole data set to such a simple equation allows us to interpret the secondary creep regime as a mere crossover from creep to crack growth.

Let us now summarize and discuss the most prominent results of this Letter. First we have shown that casein gels display a remarkable failure scenario similar to that of brittle solids and characterized by the same three successive creep regimes. Here, the power-law exponent of the primary creep [$\gp(t)\sim t^{-\alpha}$ with $\alpha=0.85\pm 0.04$] is fully accounted for by linear viscoelasticity and local measurements consistently display a homogeneous bulk deformation. Such a link between creep and viscoelasticity is shared by other biopolymer gels with power-law rheology \cite{Gobeaux:2010,Jaishankar:2013} as well as hard-sphere-like colloidal glasses \cite{Siebenburger:2012}. To better understand this link, additional creep and recovery tests were performed within the primary regime (Figs.~3 and 4 in the Supplemental Material). Successive loading/unloading sequences on a fresh sample show that (i)~the strain is not fully recovered by a few percents, (ii)~the viscoelastic properties of the material are unaltered and (iii)~strain responses are indistinguishable from one test to another. This strongly suggests that the small irreversibility observed in the primary regime mainly stems from viscous dissipation (which accounts for the nonzero viscous modulus $G''$) due to the solvent flow within the gel fibrous matrix and involves only a minute amount of local damage of the gel network. Protein gels thus appear to experience a {\it kinetically} reversible primary creep, i.e. dominated by viscous dissipation rather than by plasticity.

The logarithmic fracture growth in the tertiary creep regime constitutes our second important result. Such an evolution is also commonly reported in disordered solid materials displaying brittle rupture and interpreted in the framework of Griffith-like models based on global or local energy barriers \cite{Vanel:2009}. However these approaches all predict exponential scalings for $\tau_f(\sigma)$ while our data are best fitted by the decreasing power-law $\tau_f \sim \sigma^{-\beta}$ with $\beta=5.45\pm 0.05$ [inset of Fig.~\ref{fig1}]. This last key result suggests that thermally activated crack growth is not relevant to the present soft solid. Rather the power-law scaling is strikingly reminiscent of the Basquin law of fatigue found for a variety of heterogeneous or cellular materials under cyclic deformation \cite{Kun:2007,Basquin:1910,Kohout:2000}. Basquin law has also been recently predicted for creep experiments by FBMs that combine elastic fibers with a local yield strain and take into account damage accumulation \cite{Kun:2007,Halasz:2012}. Interestingly assuming damage accumulation to be proportional to $\sigma^\gamma$ directly leads to Basquin law with $\beta=\gamma$ and large values of $\gamma$, typically larger than 5 as found here for $\beta$, lead to macroscopic cracks due to the simultaneous rupture of a large number of fibers \cite{Halasz:2012}. More generally FBMs under elongational load predict three successive creep regimes exactly alike Fig.~\ref{fig2} with similar proportionality between $\tau_{\rm min}$ and $\tau_f$ and finite-time singularity \cite{Nechad:2005,Jagla:2011}. This highlights the relevance of FBMs in the context of creep in protein gels, whose microstructure indeed appears to be formed of strands \cite{Kalab:1983,Roefs:1990}, but also urges to study FBMs in shear geometries to check whether they would be able to predict fracture growth as observed here.

To conclude, the present time- and space-resolved study exemplifies protein gels as model, brittle-like soft solids. Prompted by the remarkable simplicity of Eq.~(\ref{e.fit}) that encompasses Andrade creep, the Monkman-Grant relation and finite-time singularity within a single equation, future modeling will undoubtedly focus on the microscopic ingredients needed to predict quantitatively the Monkman-Grant prefactor and the Basquin exponent $\beta$. The next experimental step consists in a statistical study of the fluctuations associated to crack nucleation and growth as well as a systematic investigation of other systems in order to check for universality in the irreversible creep rupture of soft solids in general and of biogels such as actin, alginate or agar gels in particular. Such a study is expected to have important implications in understanding the behavior of biomaterials under extreme stress conditions.

\begin{acknowledgments}
The authors thank M. Alava, T.~Gibaud, S.~Lindstr\"om, G.~H.~McKinley and N.~Taberlet for fruitful discussions, M.~Kal\'ab for kindly providing us with unobtainable references, and A.~Parker at Firmenich for precious advice and for providing the casein and GDL. This work was funded by the Institut Universitaire de France and by the European Research Council under the European Union's Seventh Framework Programme (FP7/2007-2013) / ERC grant agreement No.~258803. 
\end{acknowledgments}

@@

\bibitem{Caton:2008} F. Caton and C. Baravian, Rheol. Acta {\bf 47}, 601--607 (2008).

\bibitem{Divoux:2011} T. Divoux et al., Soft Matter {\bf 7}, 8409--8418 (2011).
